Bathtub Qualitative Reasoning

### 2016-10-21 paper.tex

\documentclass[a4paper]{article}
\usepackage[margin=1in]{geometry}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,amsthm,amssymb}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{pdfpages}
\usepackage{float, algorithmic, algorithm2e, program}
\usepackage{tabularx}
\usepackage{longtable}
\usepackage{epstopdf}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}[section]

@@

\newpage
\section{Algorithm}
The goal of the algorithm is to construct a state graph based for a specific model and a given starting state. We decided only to model the states that are reachable from the starting state. The reason for doing so is that we only want physically plausible states. That is we don't want disconnected states or states that cannot be reached by the system.

@@

Algorithm \ref{alg:computing_state_graph} computes the state-graph based on given an starting state. It starts with a given starting state in the processing list, then it computes all the successors of this state and adds all the successors as children to state that is being processed, as well as adding these successors to the processing list, but only if a successor was not already processed.

@@

Algorithm \ref{alg:propogating_values} propagates over all quantities and modify value based on the derivative. This is done by looping over all quantities in the model and computing all the possible values for a next state based on its current value and its derivative. For each possible next value we create a new state, thus possibly creating multiple states. Because, we can have new states created by value splits of previous quantities, we need to create new states for each state that was created previously.

@@

Algorithm \ref{alg:propogating_influences} computes the new possible derivatives based on the influences. For all new states created from the previous step (value propagation), generate a list of possible derivatives. Then if both derivatives are possible this implies ambiguity and the 0 derivative is also a possibility. If the values of the influences in the previous state did not change this means that the derivative also does not change (w.r.t.\ the influences). Finally, for each state we create multiple new states based on the possible derivatives.

\vspace{12pt}
\begin{algorithm}[H]
	TODO
	\caption{Proportionals propagation}
	\label{alg:propogating_proportionals}
\end{algorithm}
\vspace{12pt}

The last step in the algorithm is computing the new derivatives based on the proportionates. Again we loop over all the quantities in the model. Then, for each quantity we check the derivatives of the quantities it is proportional to. If a quantity it is proportional to, has itself a quantity it is proportional to, we check these quantities first, this is implemented as a recursive algorithm. If the algorithm reaches a quantity that has not proportionates or influences, we base its derivative on an exogenous influence (random). On the other hand, if the algorithm reaches a quantity that only has influences, then we know that the derivative of this quantity is already determined in the previous step (propagating the influences) and thus we do not have to go deeper into recursion. In order to prevent the algorithm from entering an infinite loop (due to cycles in the influences graph), we keep track of the quantities we visited and stop visiting them again.

When we found possible derivatives for each quantity based on its proportionates, we create new states for each new state that was created by the previous propagations (value, value constraints and influences). All the propagations steps together leave us with possible next states. Which are passed to the iterative algorithm for construction a state-graph (Algorithm~\ref{alg:computing_state_graph}).

@@

First of all we chose not to implement all the inequality constraints. The main reason for not including these is that all rules have to be coded for a general problem and thus coding them takes a discrete amount of time. Therefore we decided to implement only the most important ones, Such as the continuity constraints.

We then decided not to use higher order derivatives. However because of the generality of the algorithm, we could easily have implemented them adding, for each quantity $Q$, an another quantity $\partial Q$, and imposing constraints.
Finally we only implemented random and static derivative for the inflow, but we could have easily implemented others exogenous behaviors.

@@

In figure \ref{state-graph-base} shows all possible behaviors of the system described above. Nodes are states and arrows represent transitions from a state to every reachable states starting from that source. Nodes are labeled with the description of the current value of each quantity and its derivative.

\begin{figure}[H]\label{state-graph-base}
\centering
\includegraphics[width=\textwidth,height=7.7in,keepaspectratio]{result_base_problem.png}
\caption{Plot of the state-graph for the base problem}
\end{figure}

\section{Trace}
In order to debug the code and let users to better understand what the algorithm do we print a trace. The trace of the base problem is in \texttt{base\_trace} file and the trace of the extra problem is in \texttt{extra\_trace} file.

The trace reports how the algorithm expands the state graph and how it determines all the possible transitions for a state.

In particular the trace gives insights on the procedure for computing the possible successors. It shows how the values are assigned and how the constraints are propagated. It also tells when we discard a partial assignment because it violates value or continuity constraints.

However we decided to show the big picture of what the algorithm is doing than to print every single step in order keep the file readable. Thus we don't report all the intermediate steps and computations the programs is performing highlighting only the principal steps.

\section{Extra Problem}

\begin{itemize}
\item \textbf{Quantities:}
\begin{itemize}
\item Inflow (of water into the container) $\in [0, +]$
\item Outflow (of water out of the container) $\in [0, +, Max]$
\item Volume (of the water in the container) $\in [0, +, Max]$
\item Height (of the water column in the container) $\in [0, +, Max]$
\item Pressure (of the water column at the bottom of the container) $\in [0, +, Max]$
\end{itemize}

@@

\begin{figure}[H]\label{state-graph-extra}
	\centering
	\includegraphics[width=\textwidth,height=9.5in,keepaspectratio]{result_extra_problem.png}
	\caption{Plot of the state-graph for the extra problem}
\end{figure}

>>>>

\documentclass[a4paper]{article}
\usepackage[margin=1in]{geometry}
\usepackage[english]{babel}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,amsthm,amssymb}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{pdfpages}
\usepackage{caption}
\usepackage{float, algorithmic, algorithm2e, program}
% \usepackage{algorithm, algpseudocode}

\usepackage{tabularx}
\usepackage{longtable}
\usepackage{epstopdf}
\newtheorem{theorem}{Theorem}
\newtheorem{definition}{Definition}[section]
\newlength\myindent
\setlength\myindent{2em}
\newcommand\bindent{%
  \begingroup
  \setlength{\itemindent}{\myindent}
  \addtolength{\algorithmicindent}{\myindent}
}
\newcommand\eindent{\endgroup}

@@

\newpage
\section{Algorithm}
The goal of the algorithm is to construct a state graph based for a specific model and a given starting state. We decided only to model the states that are reachable from the starting state. The reason for doing so is that we only want physically plausible states. That is, we don't want disconnected states or states that cannot be reached by the system.

@@

Algorithm \ref{alg:computing_state_graph} computes the state-graph based on given an starting state. It starts with a given starting state in the processing list, then it computes all the successors of this state and adds all the successors as children to state that is being processed, as well as adding these successors to the processing list, but only if a successor was not already processed. After the the current state, the next state in the processing list is processed repeating this process iteratively until there are no new states in the processing list.

@@

Algorithm \ref{alg:propogating_values} propagates over all quantities and computes new values based on the derivative. For each possible next value we create a new state, thus possibly creating multiple states. Because, we can have new states created by value splits of previous quantities, we need to create new states for each state that was created by a previous split.

@@

Algorithm \ref{alg:propogating_influences} computes the new possible derivatives based on the influences. For all new states created from the previous step (value propagation), generate a list of possible derivatives. Then if both derivatives are possible this implies ambiguity and the 0 derivative is also a possibility. If the values of the influences in the previous state did not change this means that the derivative also does not change (w.r.t.\ the influences), this adds partial inequality constraints. Finally, for each state we create multiple new states based on the possible derivatives.

\vspace{12pt}
\begin{algorithm}[H]
    \For{{\bf function} prop\_proportionals(q, visited\_quantities)} {
		add q to visited\_quantities\\
		\ForAll {quantities proportionals to q} {
			\If {quantity has no influences and proportionalities} {
				return derivative based on exogenous influence
			}
			\ElseIf {quantity has no proportionalities} {
				return derivative based on derivative(s) computed in previous step
			}
			\Else{
				return prop\_proportionals(quantity)
			}
		}
	}
	\caption{Recursive function for proportionals propagation}
	\label{alg:propogating_proportionals_re}
\end{algorithm}

\vspace{12pt}
\begin{algorithm}[H]
    \For{quantity in quantities}{

			\If {quantity not in visited\_quantities} {
				derivative = prop\_proportionals(quantity)
			}

			\If{quantity in visited\_quantities does NOT violate \\epsilon ordering or invalid derivative at end values}{
				add new derivative to new\_state in new\_states
			}

    }
	\caption{Proportionals propagation}
	\label{alg:propogating_proportionals}
\end{algorithm}
\vspace{12pt}

The last step in the algorithm is computing the new derivatives based on the proportionates. Again we loop over all the quantities in the model, shown in Algorithm~\ref{alg:propogating_proportionals}. Then, for each quantity we check the derivatives of the quantities it is proportional to. If a quantity it is proportional to, has itself a quantity it is proportional to, we check these quantities first, this is implemented as a recursive algorithm, shown in Algorithm~\ref{alg:propogating_proportionals_re}. If the algorithm reaches a quantity that has not proportionates or influences, we base its derivative on an exogenous influence (random). On the other hand, if the algorithm reaches a quantity that only has influences, then we know that the derivative of this quantity is already determined in the previous step (propagating the influences) and thus we do not have to go deeper into recursion. In order to prevent the algorithm from entering an infinite loop (due to cycles in the influences graph), we keep track of the quantities we visited and stop visiting them again. Lastly the algorithm checks for both continuity of derivatives and epsilon ordering.

When we found possible derivatives for each quantity based on its proportionates, we create new states for each new state that was created by the previous propagations (value, value constraints and influences). All the propagations steps together leave us with possible next states. Which are passed to the iterative algorithm for the construction of a state-graph (Algorithm~\ref{alg:computing_state_graph}).

@@

First of all we chose not to implement all the inequality constraints. The main reason for not including these is that all rules have to be coded for a general problem and we did not find a general solution to this. Therefore we decided to implement only the most important ones, Such as the continuity constraints.

We also decided not to use higher order derivatives. However because of the generality of the algorithm, we could easily have implemented them adding, for each quantity $Q$, an another quantity $\partial Q$, and imposing constraints.
Finally we only implemented random and static derivative for the inflow, but we could have easily implemented others exogenous behaviors.

@@

Figure \ref{fig:state_graph_base} shows all possible behaviors of the system described above. Nodes are states and arrows represent transitions from a state to every reachable state. Nodes are labeled with the description of the current value of each quantity and its derivative.

\begin{figure}[H]
\centering
\includegraphics[width=\textwidth,height=7.7in,keepaspectratio]{result_base_problem.png}
\caption{Plot of the state-graph for the base problem}
\label{fig:state_graph_base}
\end{figure}

\section{Trace}
In order to debug the code as well as creating the possibility for users to better understand what the algorithm does, we generate a trace for the algorithm. The trace of the base problem is in \texttt{base\_trace.log} file and the trace of the extra problem is in \texttt{extra\_trace.log} file.

The trace reports how the algorithm expands the state graph and how it determines all the possible transitions for a state. In particular, the trace gives insights on the procedure for computing the possible successors. It shows how the values are assigned and how the constraints are propagated. It also tells when we discard a partial assignment because it violates any constraints.

\section{Extra Problem}
For the extra problem we use the following quantities and dependencies, and we get a state graph as shown in Figure~\ref{fig:state_graph_extra}.
\begin{itemize}
\item \textbf{Quantities:}
\begin{itemize}
\item Inflow (of water into the container) $\in [0, +]$
\item Outflow (of water out of the container) $\in [0, +, Max]$
\item Volume (of the water in the container) $\in [0, +, Max]$
\item Height (of the water column in the container) $\in [0, +, Max]$
\item Pressure (of the water column at the bottom of the container) $\in [0, +, Max]$
\end{itemize}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth,height=9.5in,keepaspectratio]{result_extra_problem.png}
	\caption{Plot of the state-graph for the extra problem}
	\label{fig:state_graph_extra}
\end{figure}
